Automated assessment of residual plots with computer vision models

### 2024-02-12 paper/paper.tex
update paper

\articletype{ARTICLE TEMPLATE}

@@

\author{\name{Weihao Li$^{a}$}
\affil{$^{a}$Department of Econometrics and Business Statistics, Monash
University, Clayton, VIC, Australia}
}

\thanks{CONTACT Weihao
Li. Email: \href{mailto:weihao.li@monash.edu}{\nolinkurl{weihao.li@monash.edu}}}

@@

Generating a residual plot in most statistical software is often as
straightforward as executing a line of code or clicking a button.
However, accurately interpreting a residual plot can be challenging.
Consider Figure \ref{fig:false-finding} as an example, the residuals
display a triangular shape pointing to the left. While this might
suggest heteroskedasticity, it is important to avoid over-interpreting
this visual pattern. In this case, the fitted model is correctly
specified, and the triangular shape is actually a result of the skewed
distribution of the predictors, rather than indicating a flaw in the
model.

A residual plot can exhibit various visual features, but it is crucial
to recognize that some may arise from the characteristics of predictors
and the inherent randomness of the error, rather than indicating a
violation of model assumptions \citep{li2023plot}. The concept of visual
inference, as proposed by \citet{buja2009statistical}, provides an
inferential framework to assess whether residual plots indeed contain
visual patterns inconsistent with the model assumptions. The fundamental
idea involves testing whether the actual residual plot visually differs
significantly from null plots, which are created using residuals
generated from the null distribution. Typically, this is accomplished
through the lineup protocol. In this approach, the real residual plot is
embedded within a lineup alongside several null plots. If the real
residual plot can be distinguished from the lineup, it provides evidence
for rejecting the null hypothesis.

The practice of delivering a residual plot as a lineup is generally
regarded as a valuable approach. Beyond its application in residual
diagnostics, the lineup protocol has integrated into the analysis of
diverse subjects. For instance,
\cite{loy2013diagnostic, loy2014hlmdiag, loy2015you} illustrate its
applicability in diagnosing hierarchical linear models. Additionally,
\citet{widen2016graphical} demonstrates its utility in geographical
research, while \citet{krishnan2021hierarchical} explores its
effectiveness in forensic examinations.

@@

The utilization of computers to interpret data plots has a rich history,
with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
focusing on scatterplot diagnostics. \citet{wilkinson2005graph} expanded
on this work, introducing graph theoretic scagnostics, which encompassed
nine computable measures applied to planar proximity graphs. These
measures, including, but not limited to, ``Outlying'', ``Skinny'',
``Stringy'', ``Straight'', ``Monotonic'', ``Skewed'', ``Clumpy'', and
``Striated'' aimed to characterize outliers, shape, density, trend,
coherence and other characteristics of the data. While this approach has
been inspiring, there is a recognition \citep{buja2009statistical} that
it may not capture all the necessary visual features that differentiate
actual residual plots from null plots. A more promising alternative
entails enabling computers to learn the function for extracting visual
features from residual plots. Essentially, this means empowering
computers to discern the crucial visual features for residual
diagnostics and determining the method to extract them. Modern computer
vision models are well-suited for addressing this challenge.

Modern computer vision models often rely on deep neural networks with
convolutional layers \citep{fukushima1982neocognitron}. These layers
leverage hierarchical patterns in data, downsizing and transforming
images by summarizing information in a small space. Numerous studies
have demonstrated the efficacy of convolutional layers in addressing
various vision tasks, including image recognition \citep{rawat2017deep}.
Despite the widespread use of computer vision models in fields like
computer-aided diagnosis \citep{lee2015image}, pedestrian detection
\citep{brunetti2018computer}, and facial recognition
\citep{emami2012facial}, their application in reading data plots remains
limited. While some studies have explored the use of computer vision
models for tasks such as reading recurrence plots for time series
regression \citep{ojeda2020multivariate}, time series classification
\citep{chu2019automatic, hailesilassie2019financial, hatami2018classification, zhang2020encoding},
anomaly detection \citep{chen2020convolutional}, and pairwise causality
analysis \citep{singh2017deep}, the application of reading residual
plots with computer vision models represents a relatively new field of
study.

@@

@@

\hypertarget{different-input-formats}{%
\subsubsection{Different input formats}\label{different-input-formats}}

Deep learning models are in general very sensitive to the input data.
The quality and relevance of the input data greatly influence the
model's capacity to generate insightful and meaningful results.

A straightforward approach involves feeding a vector of residuals along
with a vector of fitted values, essentially providing all the necessary
information for creating a residuals vs fitted values plot. However, a
drawback of this method is the dynamic input size, which changes based
on the number of observations. For modern computer vision models
implemented in mainstream software like TensorFlow
\citep{abadi2016tensorflow}, the input shape is typically fixed. One
solution is to pad the input vectors with leading or trailing zeros when
the input tensor expects longer vectors, but it may fail if the input
vector surpasses the designed length. Another strategy is to summarize
the residuals and fitted values separately using histograms and utilize
the counts as the input. By controlling the number of bins in the
histograms, it becomes possible to provide fixed-length input vectors.
Still, since the histogram only captures the marginal distribution of
residuals and fitted values respectively, it can not be used to
differentiate visual patterns with same marginal distributions but
different joint distributions.

Another approach is to use an image as input. The primary advantage of
using an image, as opposed to the vector format, is the utilization of
the sophisticated image processing architectures developed over the
years, such as the VGG16 architecture proposed in
\citet{simonyan2014very}. These architectures can effectively capture
and summarize spatial information from nearby pixels, which is less
straightforward with vector input. The only considerations are the image
resolution and the aesthetics of the residual plot. In general, higher
resolution provides more information to the model but comes with the
trade-off of increased complexity and greater difficulty in training. As
for the aesthetics of the residual plot, a practical solution is to
consistently present residual plots in the same style to the model. This
implies that the model can not accept arbitrary images as input but
requires the use of the same preprocessing software to convert provided
residuals and fitted values into a standardized style for the residual
plot.

Providing multiple residual plots to the model, such as a pair of plots,
a triplet or a lineup is also a possible option.
\citet{chopra2005learning} has shown that computer vision models
designed for image comparison can assess whether a pair of images are
similar or dissimilar. Applied to our specific problem, we can define
null plots of a fitted model to be similar to each other, while
considering actual residual plots to be distinct from null plots of any
fitted model. A triplet constitutes a set of three images, denoted as
\(image_1\), \(image_2\) and \(image_3\). It is often used to predict
whether \(image_2\) or \(image_3\) is more similar to \(image_1\),
proving particularly useful for establishing rankings between samples.
For this setup, we can apply the same criteria to define similarity
between images. However, it's important to note that these two
approaches usually require additional considerations regarding the loss
function and, at times, non-standard training processes due to shared
weights between different convolutional blocks.

Presenting a lineup to a model and asking it to predict which residual
plot is the most different one aligns closely with the lineup protocol.
However, if a lineup consists of a large amount of residual plots, the
resolution of the input image could become quite large, posing
challenges in training the model. This approach was experimented in a
pilot study conducted by us but the performance of the trained model was
sub-optimal.

@@

\hypertarget{different-output-formats}{%
\subsubsection{Different output
formats}\label{different-output-formats}}

Given that the input is a single residual plot represented as a
fixed-resolution image, the output from the computer vision model can
take one of two formats: binary or numeric. This choice determines
whether the model belongs to a classification model or a regression
model. The binary outcome could be designed to indicate whether the
input image is a null plot, or whether the input image would be rejected
in a visual test conducted by humans. The latter option requires data
from prior human subject experiments, presenting difficulties in
controlling the quality of data due to variations in experimental
settings across different studies. Additionally, some visual inference
experiments are unrelated to linear regression models and residual plot
diagnostics, resulting in a limited amount of available training data.

Alternatively, the output could be designed as a meaningful numeric
measure, such as the average distance to good residual plots. This
approach necessitates defining a distance measure between residual
plots, which may be a non-trivial task. However, once a meaningful
distance measure is established, the prediction from the computer vision
model becomes an interpretable value with tangible significance. Studies
have demonstrated that defining a proper distance between images can
enhance the matching accuracy in image search compared to a binary
outcome model (ref here).

@@

In a visual test, the observer will be asked to choose one or more plots
that stand out as most distinct from others in a given lineup. To
develop a computer vision model for assessing residual plots within the
visual inference framework, it is important to precisely define a
numerical measure of ``difference'' or ``distance'' between plots. This
distance can take the form of a basic statistical operation on pixels,
such as the sum of square differences. Alternatively, it could involve
established image similarity metrics like the Structural Similarity
Index Measure \citep{wang2004image}. The challenge lies in the fact that
metrics tailored for image comparison may not be suitable for evaluating
data plots, where only essential plot elements require assessment
\citep{chowdhury2018measuring}. Furthermore, scagnostics mentioned in
Section \ref{introduction} could be used to construct distance metrics
for data plots comparison, but the functional form still needs to be
carefully refined to accurately reflect the extent of the violations.

@@

The distance metrics proposed in this paper takes into account the fact
that we try to measure how different a residual plot is from a good
residual plot, or in other words, how different a given fitted model is
from a correctly specified model. For the classical normal linear
regression model, residuals \(\boldsymbol{e}\) are derived from the
fitted values \(\hat{\boldsymbol{y}}\) and observed values
\(\boldsymbol{y}\). Suppose the data generating process is known and the
regression model is correctly specified, by the Frisch-Waugh-Lowell
theorem \citep{frisch1933partial}, residuals \(\boldsymbol{e}\) can also
be treated as random variables and written as a linear transformation of
the error \(\boldsymbol{\varepsilon}\) formulated as
\(\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon}\), where
\(\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the residual operator and an idempotent matrix, \(\boldsymbol{X}\) is
the design matrix, \(\boldsymbol{I}_n\) is a \(n\) by \(n\) identity
matrix, and \(n\) is the number of observations.

One of the assumptions of the classical normal linear regression model
is the error \(\boldsymbol{\varepsilon}\) follows a multivariate normal
distribution with zero mean and constant variance, i.e.,
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\).
It can be known that residuals \(\boldsymbol{e}\) also follow a certain
probability distribution transformed from the multivariate normal
distribution, which will be denoted as \(Q\). This reference
distribution \(Q\) summarizes what good residuals should follow given
the predictors are known and fixed.

In ordinary least square, the minimization of the sum of square
residuals implies \(\sum_{i=1}^{n} e_i = 0\), making any residual value
to be a linear combination of the remaining \(n - 1\) residuals. This
effectively means \(\text{rank}(\boldsymbol{R}) = n - 1 < n\) and \(Q\)
is a degenerate multivariate distribution. To capture the
characteristics of \(Q\), such as moments, we can simulate a large
numbers of \(\boldsymbol{\varepsilon}\) and transform it to
\(\boldsymbol{e}\) to get the empirical estimates. For simplicity, we
replaced the variance-covariance matrix of residuals
\(\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}' = \boldsymbol{R}\sigma^2\)
with a full-rank diagonal matrix
\(\text{diag}(\boldsymbol{R}\sigma^2)\), where \(\text{diag}(.)\) sets
the non-diagonal entries of a matrix to zeros. The resulting
distribution for \(\boldsymbol{Q}\) is
\(N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))\).

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different from \(Q\). For example,
if the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but not included in \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different from \(Q\)
because the residual operator obtained from the fitted model will not be
the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate lognormal distribution, \(P\) will usually be skewed and
has a long tail.

@@

Define a proper distance between distributions is usually easier than
define a proper distance between data plots. Given the actual residual
distribution \(Q\) and the reference residual distribution \(P\), we
used a distance metric based on Kullback-Leibler divergence
\citep{kullback1951information} to quantify the difference between two
distributions

@@

This distance metric was first proposed in \citet{li2023plot}. It was
mainly designed for measuring the effect size of non-linearity and
heteroskedasticity in a residual plot. \citet{li2023plot} showed that,
for a classical normal linear regression model that omits a necessary
higher-order predictors \(\boldsymbol{Z}\), and incorrectly assumes
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\)
while in fact
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})\)
with \(\boldsymbol{V}\) being an arbitrary symmetric positive
semi-definite matrix, \(Q\) can be represented as
\(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))\).
Note that the variance-covariance matrix is replaced with the diagonal
matrix to ensure it is a full-rank matrix.

@@

For non-normal error \(\boldsymbol{\varepsilon}\), the actual residual
distribution \(P\) is unlikely to be a multivariate normal distribution.
Thus, equation \ref{eq:kl-2} given in \citet{li2023plot} will not be
applicable to models with non-normality violations.

@@

The algorithm for approximating equation \ref{eq:kl-1} starts from
simulating \(m\) sets of \(\boldsymbol{\varepsilon}\) according to the
error distribution. The simulated errors are stored in a matrix
\(\boldsymbol{A}\) with \(n\) rows and \(m\) columns, where each column
of \(\boldsymbol{A}\) is a set of realized values of
\(\boldsymbol{\varepsilon}\). Then, we can get \(m\) sets of realized
values of \(\boldsymbol{e}\) stored in the matrix \(\boldsymbol{B}\) by
applying the residual operator
\(\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}\). Furthermore, kernel
density estimation (KDE) with Gaussian kernel and optimal bandwidth
selected by the Silverman's rule of thumb \citep{silverman2018density}
is applied on each row of \(B\) to estimate \(p_i(e_i)\) for
\(i = 1, ..., n\). The KDE computation is done by the \texttt{density}
function in R.

@@

\hypertarget{approximation-of-the-distance-metric}{%
\subsubsection{Approximation of the distance
metric}\label{approximation-of-the-distance-metric}}

In the previous sections, we have defined a distance metric given in
equation \ref{eq:kl-0} for quantifying the difference between the actual
residual distribution \(P\) and an ideal reference distribution \(Q\).
You may have noticed that this distance metric can only be computed when
the data generating process is known. In reality, we often have no
knowledge about the data generating process, otherwise we do not need to
fit a linear regression model in the first place.

What we proposed in this paper is a method to approximate this distance
with a residual plot. Let \(D\) be the result of equation \ref{eq:kl-0}
indicating the extent of the model violations, and our estimator
\(\hat{D}\) is formulated as

@@

With the approximated distance \(\hat{D}\), we will be able to know how
different the underlying distribution of the residuals is from a good
residual distribution. This is meaningful way to check if a residual
plot is a good residual plot, and to know the strength of the visual
signal embedded in the residual plot.

The approximated distance \(\hat{D}\) is not expected to be the same as
the original distance \(D\). This is largely because information
contained in a single residual plot is limited and it may not be able to
summarise all the characteristics of the residual distribution. For a
given residual distribution \(P\), we can generate many different
residual plots. Some of them share similar visual patterns, but some of
them could be visually very different from the rest, especially for
models with small \(n\). This suggests the error of the approximation
will vary depends on whether the observed residual plot is
representative or not.

@@

This property is not an issue if \(\hat{D} \gg 0\) for which the visual
signal of the residual plot is very strong, and we usually do not need
any further examination of the significance of the result. However, if
the visual pattern is weak or moderate, having \(\hat{D}\) will not be
sufficient to determine if \(H_0\) that the model is correctly specified
should be rejected.

To solve this issue while aligning with the principle of visual
inference, the approximated distance \(\hat{D}\) can be viewed as a test
statistic. And the null distribution of this statistic can be
approximated by the empirical distribution of \(\hat{D}_{null}\) for a
large amount of null plots of a fitted model. The procedure involves
applying the residual rotation technique \citep{buja2009statistical} on
the fitted model to obtain null residuals. The null residuals are then
used to make null plots and fed into the model to get predictions. The
empirical distribution is constructed with the predicted distance.

@@

While observational data is frequently employed in training models for
real-world applications, the data generating process of observational
data often remains unknown, making computation for our target variable
\(D\) unattainable. Consequently, the computer vision models developed
in this study were trained using synthetic data. This approach provides
us with precise label annotations. Additionally, it ensures a large and
diverse training dataset, as we have control over the data generating
process, and the simulation of the training data is relatively
cost-effective.

@@

\begin{itemize}
\tightlist
\item
  a diagram for the model
\end{itemize}

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/rd-human-1}

\caption{Residuals vs distance plot on lineups used in the human subject experiment. The blue line is a smoothing curve produced by fitting a gnealized additive model.}\label{fig:rd-human}
\end{figure}

>>>>

\articletype{}

@@

\author{\name{Weihao Li$^{a}$, Dianne Cook$^{a}$, Emi Tanaka$^{b,
c}$, Susan VanderPlas$^{d}$, Klaus Ackermann$^{a}$}
\affil{$^{a}$Department of Econometrics and Business Statistics, Monash
University, Clayton, VIC, Australia; $^{b}$Biological Data Science
Institute, Australian National University, Acton, ACT,
Australia; $^{c}$Research School of Finance, Actuarial Studies and
Statistics, Australian National University, Acton, ACT,
Australia; $^{d}$Department of Statistics, University of Nebraska,
Lincoln, Nebraska, USA}
}

\thanks{CONTACT Weihao
Li. Email: \href{mailto:weihao.li@monash.edu}{\nolinkurl{weihao.li@monash.edu}}, Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}, Emi
Tanaka. Email: \href{mailto:emi.tanaka@anu.edu.au}{\nolinkurl{emi.tanaka@anu.edu.au}}, Susan
VanderPlas. Email: \href{mailto:susan.vanderplas@unl.edu}{\nolinkurl{susan.vanderplas@unl.edu}}, Klaus
Ackermann. Email: \href{mailto:Klaus.Ackermann@monash.edu}{\nolinkurl{Klaus.Ackermann@monash.edu}}}

@@

Generating a residual plot in most statistical software is often as
straightforward as executing a line of code or clicking a button.
However, accurately interpreting a residual plot can be challenging.
Consider Figure \ref{fig:false-finding} as an example, the residuals
display a triangular shape pointing to the left. While this might
suggest heteroskedasticity, it is important to avoid over-interpreting
this visual pattern. In this case, the fitted regression model is
correctly specified, and the triangular shape is actually a result of
the skewed distribution of the predictors, rather than indicating a flaw
in the model.

A residual plot can exhibit various visual features, but it is crucial
to recognize that some may arise from the characteristics of predictors
and the inherent randomness of the error, rather than indicating a
violation of model assumptions \citep{li2023plot}. The concept of visual
inference, as proposed by \citet{buja2009statistical}, provides an
inferential framework to assess whether residual plots indeed contain
visual patterns inconsistent with the model assumptions. The fundamental
idea involves testing whether the actual residual plot visually differs
significantly from null plots, where null plots are plotted with
residuals generated from the residual rotation distribution
\citep{langsrud2005rotation}, which is a distribution consistent with
the null hypothesis \(H_0\) that the linear regression model is
correctly specified. Typically, the visual test is accomplished through
the lineup protocol, where the real residual plot is embedded within a
lineup alongside several null plots. If the real residual plot can be
distinguished from the lineup, it provides evidence for rejecting
\(H_0\).

The practice of delivering a residual plot as a lineup is generally
regarded as a valuable approach. Beyond its application in residual
diagnostics, the lineup protocol has integrated into the analysis of
diverse subjects. For instance,
\cite{loy2013diagnostic, loy2014hlmdiag, loy2015you} illustrated its
applicability in diagnosing hierarchical linear models. Additionally,
\citet{widen2016graphical} demonstrated its utility in geographical
research, while \citet{krishnan2021hierarchical} explored its
effectiveness in forensic examinations.

@@

The utilization of computers to interpret data plots has a rich history,
with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
focusing on scatterplot diagnostics. \citet{wilkinson2005graph} expanded
on this work, introducing graph theoretic scagnostics, which encompassed
computable measures applied to planar proximity graphs. These measures,
including, but not limited to, ``Outlying'', ``Skinny'', ``Stringy'',
``Straight'', ``Monotonic'', ``Skewed'', ``Clumpy'', and ``Striated''
aimed to characterize outliers, shape, density, trend, coherence and
other characteristics of the data. While this approach has been
inspiring, there is a recognition \citep{buja2009statistical} that it
may not capture all the necessary visual features that differentiate
actual residual plots from null plots. A more promising alternative
entails enabling machines to learn the function for extracting visual
features from residual plots. Essentially, this means empowering
computers to discern the crucial visual features for residual
diagnostics and determining the method to extract them.

Modern computer vision models are well-suited for addressing this
challenge. They rely on deep neural networks with convolutional layers
\citep{fukushima1982neocognitron}. These layers leverage hierarchical
patterns in data, downsizing and transforming images by summarizing
information in a small space. Numerous studies have demonstrated the
efficacy of convolutional layers in addressing various vision tasks,
including image recognition \citep{rawat2017deep}. Despite the
widespread use of computer vision models in fields like computer-aided
diagnosis \citep{lee2015image}, pedestrian detection
\citep{brunetti2018computer}, and facial recognition
\citep{emami2012facial}, their application in reading data plots remains
limited. While some studies have explored the use of computer vision
models for tasks such as reading recurrence plots for time series
regression \citep{ojeda2020multivariate}, time series classification
\citep{chu2019automatic, hailesilassie2019financial, hatami2018classification, zhang2020encoding},
anomaly detection \citep{chen2020convolutional}, and pairwise causality
analysis \citep{singh2017deep}, the application of reading residual
plots with computer vision models represents a relatively new field of
study.

@@

\begin{figure}[!h]

@@

\hypertarget{input-formats}{%
\subsubsection{Input formats}\label{input-formats}}

Deep learning models are in general very sensitive to the input data.
The quality and relevance of the input data greatly influence the
model's capacity to generate insightful and meaningful results. There
are several designs of the input format can be considered.

A straightforward design involves feeding a vector of residuals along
with a vector of fitted values, essentially providing all the necessary
information for creating a residuals vs fitted values plot. However, a
drawback of this method is the dynamic input size, which changes based
on the number of observations. For modern computer vision models
implemented in mainstream software like TensorFlow
\citep{abadi2016tensorflow}, the input shape is typically fixed. One
solution is to pad the input vectors with leading or trailing zeros when
the input tensor expects longer vectors, but it may fail if the input
vector surpasses the designed length. Another strategy is to summarize
the residuals and fitted values separately using histograms and utilize
the counts as the input. By controlling the number of bins in the
histograms, it becomes possible to provide fixed-length input vectors.
Still, since histograms only capture the marginal distribution of
residuals and fitted values respectively, they can not be used to
differentiate visual patterns with same marginal distributions but
different joint distributions.

Another design involves using an image as input. The primary advantage
of this design, as opposed to the vector format, is the availability of
the existing and sophisticated image processing architectures developed
over the years, such as the VGG16 architecture proposed in
\citet{simonyan2014very}. These architectures can effectively capture
and summarize spatial information from nearby pixels, which is less
straightforward with vector input. The main considerations are the image
resolution and the aesthetics of the residual plot. In general, higher
resolution provides more information to the model but comes with the
trade-off of increased complexity and greater difficulty in training. As
for the aesthetics of the residual plot, a practical solution is to
consistently present residual plots in the same style to the model. This
implies that the model can not accept arbitrary images as input but
requires the use of the same preprocessing pipeline to convert residuals
and fitted values into a standardized-style residual plot.

Providing multiple residual plots to the model, such as a pair of plots,
a triplet or a lineup is also a possible option.
\citet{chopra2005learning} have shown that computer vision models
designed for image comparison can assess whether a pair of images are
similar or dissimilar. Applied to our specific problem, we can define
null plots of a fitted regression model to be similar to each other,
while considering actual residual plots to be distinct from null plots
of any fitted regression model. A triplet constitutes a set of three
images, denoted as \(image_1\), \(image_2\) and \(image_3\). It is often
used to predict whether \(image_2\) or \(image_3\) is more similar to
\(image_1\), proving particularly useful for establishing rankings
between samples. For this setup, we can apply the same criteria to
define similarity between images. However, it is important to note that
these two approaches usually require additional considerations regarding
the loss function and, at times, non-standard training processes due to
shared weights between different convolutional blocks.

Presenting a lineup to a model aligns closely with the lineup protocol.
However, as the number of residual plots in a lineup increases, the
resolution of the input image grows rapidly, posing challenges in
training the model. We experimented with this approach in a pilot study,
but the performance of the trained model was sub-optimal.

@@

\hypertarget{output-formats}{%
\subsubsection{Output formats}\label{output-formats}}

Given that the input is a single residual plot represented as a
fixed-resolution image, the output from the computer vision model can
take one of two forms: binary or numeric. This choice determines whether
the model belongs to a classification model or a regression model. The
binary outcome encoded as \(0\) and \(1\) could be used to represent
whether the input image is a null plot, or whether the input image would
be rejected in a visual test conducted by humans. Training a model
following the latter option requires data from prior human subject
experiments, presenting difficulties in controlling the quality of data
due to variations in experimental settings across different studies.
Additionally, some visual inference experiments are unrelated to linear
regression models or residual plot diagnostics, resulting in a limited
amount of available training data.

Alternatively, the output could be a meaningful and interpretable
numerical measure useful for assessing residual plots, such as the
strength of suspicious visual patterns reflecting the extent of model
violations and the difficulty index for identifying whether a residual
plot has no issues. However, these numeric measures are often informally
used in daily communication but are not typically formalized or
rigorously defined. For the purpose of training a model, this numeric
measure has to be quantifiable.

In this study, we chose to define and use a distance measure to quantify
the difference between the residual plot and the null plots.
\citet{vo2016localizing} have also demonstrated that defining a proper
distance between images can enhance the matching accuracy in image
search compared to a binary outcome model.

@@

In a visual test, the observer will be asked to choose one or more plots
that stand out as most distinct from others in a given lineup. To
develop a computer vision model for assessing residual plots within the
visual inference framework, it is important to precisely define a
numerical measure of ``difference'' or ``distance'' between plots. This
distance can take the form of a basic statistical operation on pixels,
such as the sum of square differences. Alternatively, it could involve
established image similarity metrics like the Structural Similarity
Index Measure \citep{wang2004image}. The challenge lies in the fact that
metrics tailored for image comparison may not be suitable for evaluating
data plots, where only essential plot elements require assessment
\citep{chowdhury2018measuring}. Furthermore, scagnostics mentioned in
Section \ref{introduction} could be used to construct distance metrics
for residual plots comparison, but the functional form still needs to be
carefully refined to accurately reflect the extent of the violations.

@@

The distance measure proposed in this study takes into account the fact
that we tried to measure how different a residual plot is from a good
residual plot, or in other words, how different a given fitted
regression model is from a correctly specified model. For the classical
normal linear regression model, residuals \(\boldsymbol{e}\) are derived
from the fitted values \(\hat{\boldsymbol{y}}\) and observed values
\(\boldsymbol{y}\). Suppose the data generating process is known and the
regression model is correctly specified, by the Frisch-Waugh-Lowell
theorem \citep{frisch1933partial}, residuals \(\boldsymbol{e}\) can also
be treated as random variables and written as a linear transformation of
the error \(\boldsymbol{\varepsilon}\) formulated as
\(\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon}\), where
\(\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the residual operator and an idempotent matrix, \(\boldsymbol{X}\) is
the design matrix, \(\boldsymbol{I}_n\) is a \(n\) by \(n\) identity
matrix, and \(n\) is the number of observations.

One of the assumptions of the classical normal linear regression model
is the error \(\boldsymbol{\varepsilon}\) follows a multivariate normal
distribution with zero mean and constant variance, i.e.,
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\).
It can be known that residuals \(\boldsymbol{e}\) also follow a certain
probability distribution transformed from the multivariate normal
distribution, which will be denoted as \(Q\). This reference
distribution \(Q\) summarizes what good residuals should follow given
the design matrix \(\boldsymbol{X}\) is known and fixed.

In ordinary least square, the minimization of the sum of square
residuals implies \(\sum_{i=1}^{n} e_i = 0\), making any residual value
to be a linear combination of the remaining \(n - 1\) residuals. This
effectively means \(\text{rank}(\boldsymbol{R}) = n - 1 < n\) and \(Q\)
is a degenerate multivariate distribution. To capture the
characteristics of \(Q\), such as moments, we can simulate a large
numbers of \(\boldsymbol{\varepsilon}\) and transform it to
\(\boldsymbol{e}\) to get the empirical estimates. For simplicity, in
this study, we replaced the variance-covariance matrix of residuals
\(\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}' = \boldsymbol{R}\sigma^2\)
with a full-rank diagonal matrix
\(\text{diag}(\boldsymbol{R}\sigma^2)\), where \(\text{diag}(.)\) sets
the non-diagonal entries of a matrix to zeros. The resulting
distribution for \(\boldsymbol{Q}\) is
\(N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))\).

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different from \(Q\). For example,
if the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but not included in \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different from \(Q\)
because the residual operator obtained from the fitted regression model
will not be the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate lognormal distribution, \(P\) will usually be skewed and
has a long tail.

@@

Define a proper distance between distributions is usually easier than
define a proper distance between data plots. Given the actual residual
distribution \(Q\) and the reference residual distribution \(P\), we
used a distance measure based on Kullback-Leibler divergence
\citep{kullback1951information} to quantify the difference between two
distributions

@@

This distance measure was first proposed in \citet{li2023plot}. It was
mainly designed for measuring the effect size of non-linearity and
heteroskedasticity in a residual plot. \citet{li2023plot} have showed
that, for a classical normal linear regression model that omits a
necessary higher-order predictors \(\boldsymbol{Z}\), and incorrectly
assumes
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\)
while in fact
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})\)
with \(\boldsymbol{V}\) being an arbitrary symmetric positive
semi-definite matrix, \(Q\) can be represented as
\(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))\).
Note that the variance-covariance matrix is replaced with the diagonal
matrix to ensure it is a full-rank matrix.

@@

For non-normal error \(\boldsymbol{\varepsilon}\), the actual residual
distribution \(P\) is unlikely to be a multivariate normal distribution.
Thus, equation \ref{eq:kl-2} given in \citet{li2023plot} will not be
applicable to models violating the normality assumption.

@@

Given \(\boldsymbol{X}\) and \(\boldsymbol{\beta}\), the algorithm for
approximating equation \ref{eq:kl-1} starts from simulating \(m\) sets
of observed values \(\boldsymbol{y}\) according to the data generating
process. The observed values are stored in a matrix \(\boldsymbol{A}\)
with \(n\) rows and \(m\) columns, where each column of
\(\boldsymbol{A}\) is a set of observed values. Then, we can get \(m\)
sets of realized values of \(\boldsymbol{e}\) stored in the matrix
\(\boldsymbol{B}\) by applying the residual operator
\(\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}\). Furthermore, kernel
density estimation (KDE) with Gaussian kernel and optimal bandwidth
selected by the Silverman's rule of thumb \citep{silverman2018density}
is applied on each row of \(B\) to estimate \(p_i(e_i)\) for
\(i = 1, ..., n\). The KDE computation can be done by the
\texttt{density} function in R.

@@

\hypertarget{approximation-of-the-distance-measure}{%
\subsubsection{Approximation of the distance
measure}\label{approximation-of-the-distance-measure}}

In the previous sections, we have defined a distance measure given in
equation \ref{eq:kl-0} for quantifying the difference between the actual
residual distribution \(P\) and an ideal reference distribution \(Q\).
You may have noticed that this distance measure can only be computed
when the data generating process is known. In reality, we often have no
knowledge about the data generating process, otherwise we do not need to
fit a linear regression model in the first place.

We tried to train a computer vision model to approximate this distance
measure with a residual plot. Let \(D\) be the result of equation
\ref{eq:kl-0} indicating the extent of the model violations, and our
estimator \(\hat{D}\) is formulated as

@@

With the approximated distance \(\hat{D}\), we will be able to know how
different the underlying distribution of the residuals is from a good
residual distribution. It also provides information for the strength of
the visual signal embedded in the residual plot.

The approximated distance \(\hat{D}\) is not expected to be the same as
the original distance \(D\). This is largely because information
contained in a single residual plot is limited and it may not be able to
summarise all the important characteristics of the residual
distribution. For a given residual distribution \(P\), we can generate
many different residual plots. Some of them share similar visual
patterns, but some of them could be visually very different from the
rest, especially for regression models with small \(n\). This suggests
the error of the approximation will vary depends on whether the observed
residual plot is representative or not.

@@

This property is not an issue if \(\hat{D} \gg 0\) for which the visual
signal of the residual plot is very strong, and we usually do not need
any further examination of the significance of the result. However, if
the visual pattern is weak or moderate, having \(\hat{D}\) will not be
sufficient to determine if \(H_0\) should be rejected.

To solve this issue while aligning with the principle of visual
inference, \(\hat{D}\) can be viewed as a test statistic. And the null
distribution of this statistic can be approximated by the empirical
distribution of \(\hat{D}_{null}\), where \(\hat{D}_{null}\) is the
approximated distance for a null plot simulated from the fitted
regression model. The approximation of the distribution involves
applying the residual rotation technique \citep{buja2009statistical} on
the fitted regression model to obtain null residuals. The null residuals
are then used to make null plots and fed into the computer vision model
to get predictions. The empirical distribution is constructed with the
approximated distance.

@@

While observational data is frequently employed in training models for
real-world applications, the data generating process of observational
data often remains unknown, making computation for our target variable
\(D\) unattainable. Consequently, the computer vision models developed
in this study were trained using synthetic data. This approach provided
us with precise label annotations. Additionally, it ensured a large and
diverse training dataset, as we had control over the data generating
process, and the simulation of the training data was relatively
cost-effective.

@@

@@
